Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-04-22 paper.tex
update

%NEW
\newcommand{\rmn}{\mathrm}
\newcommand{\bx}{{\bf x}}
\newcommand{\fg}{{F_\gamma}}
\newcommand{\sg}{{S_\gamma}}
\newcommand{\psf}{\theta_\rmn{res}}
\newcommand{\ph}{\rmn{ph}}
\newcommand{\eph}{E_\ph}
\newcommand{\nph}{N_\ph}
\newcommand{\vir}{\rmn{vir}}
\newcommand{\gal}{\rmn{gal}}
\newcommand{\sd}{\rmn{SD}}
\newcommand{\clu}{\rmn{cl}}
\newcommand{\sfe}{\rmn{sfe}}
\newcommand{\sub}{\rmn{sub}}
\newcommand{\msun}{M_\odot}
\newcommand{\ee}{E_\rmn{e}}
\newcommand{\stars}{\rmn{stars}}
\newcommand{\dust}{\rmn{dust}}
\newcommand{\lx}{L_\rmn{X}}
\newcommand{\s}{\rmn{s}}
\newcommand{\Kp}{\rmn{K}^\prime}
\newcommand{\Ip}{\rmn{I}^\prime}
\newcommand{\Js}{\rmn{J}^*}
\newcommand{\Jp}{\rmn{J}^\prime}
\newcommand{\B}{\rmn{B}}
\newcommand{\bsub}{\B_\rmn{sub}}
\newcommand{\qCR}{q_{\rmn{CR}-\ensuremath{\pi^0}}}
\newcommand{\ds}{{\sc DarkSUSY}}

@@

\begin{abstract}
... to be written...
\end{abstract}

@@

% --- section:  --- %
\section{Theory}
\label{sect:theory}

@@

\frac{\dd \sg}{\dd \eg}(\psi,\eg) = \int_{\Delta\Omega} \dd\Omega \int_{\rm los}

@@

radiation (BM-Cont), and CR proton induced $\pi^0$:s decaying into

@@

is given by $\B_{\rmn{tot},i}(r,\sigv) =
\B_\sfe(\sigv)\,\B_{\sub,i}(r)$. It is the product of enhancement factors
from SFE $\B_\sfe(\sigv)$ (see Sect.~\ref{sect:LP}) and from
substructure enhancement over the smooth halo contribution
$\B_{\sub,i}(r) = 1+q_\sub(r)/q_{\rmn{sm},i}(r)$ (see
Sect.~\ref{sect:subst}). The DM source function from the smooth halo
for each process is written on the form:

@@

Here the subscript $j$ runs over all kinematically allowed gamma-ray
producing channels each with the spectrum $\frac{\dd

@@

profile provide a good fit to both the observed and simulated

@@

should also be noted that recent detailed galaxy cluster observations
find a core like density profile with an inner slope of $\beta=0.6$,
where $\beta\gtrsim 1$ can be ruled out to a $>95$\% confidence
\cite{2011ApJ...728L..39N}. In Eq.~(\ref{eq:dens_ein}) we denote the
density where the profile has a slope of $-2$ by $\rho_{-2}$, and the
radius by $r_{-2}$. Assuming that all the flux from an NFW profile

@@

that $\rho_{-2} = \rhos/4$.

@@

\subsubsection{Substructures}
\label{sect:subst}
High-resolution DM only simulations of Milky Way (MW) type halos find
substantial amount of substructures in the periphery of DM halos,
while the substructures in the center are erased due to dynamical
disruption. Since the rate of which DM is annihilating depends on the
density squared, the resulting flux from substructures is boosted
compared to the smooth density distribution. Recent high resolution
simulations suggest a flux enhancement from substructures of the order
of ten up to a few 100 \cite{2008MNRAS.391.1685S, 2008JPhCS.125a2008K}
for a MW size halo. We use a double power-law function to
fit \footnote{Our approach of fitting the scaling behavior of
  $\rmn{L_{sub}}(<r)$ directly from numerical simulations
  self-consistently accounts for the radial dependence of the
  substructure concentration \cite{2008MNRAS.391.1685S}.} the
luminosity from the smooth component of substructures
(i.e. substructures within substructures are not included) inside
radius $r$ which is determined for the Aquarius simulations
\cite{2008MNRAS.391.1685S,2008Natur.456...73S}. Our best fit is given
by
\begin{eqnarray}
  {\rm L_{sub}}(<r) &=& 0.76\,C_1\,C_2(\mvir)\,{\rm L_{200sm}}\,x^{a_0
    \displaystyle{x}^{a_1}}\quad\rmn{where} \nonumber\\
  a_0 = 0.95\,,\quad  a_1 = -0.27 \qquad \rmn{and} \nonumber\\
  C(\mvir) = C_1\left(\frac{\mvir}{M_{\rm lim}}\right)^{0.226}\,.
C_1  
\label{eq:LSub}
\end{eqnarray}
Here

where ${\rm L_{200sm}}$ is the luminosity from the smooth halo without
substructures within $\rvir$ and $x=
\left(\frac{r}{\rvir}\right)$. The first normalization constant is
derived from the simulations:
\begin{equation}
  C_1=\left(\frac{M_{\rm res,sim}}{M_{\rm lim}}\right)^{0.226}\,,
   \label{eq:C1}
\end{equation}
where $M_{\rm res,sim}=10^5\msun$ is the mass of the smallest resolved
subhalos in the MW simulation, and $M_{\rm lim}$ is the theoretical
smallest mass of subhalos determined from free streaming length of DM
(CHECK). Here we assume that the power-law scaling relation in
Eq.~(\ref{eq:C1}) is valid down to the free streaming mass of DM
halos, which is conventionally in the cold dark matter universe
\cite{2001PhRvD..64h3507H,2005JCAP...08..003G} taken to be
$10^{-6}\msun$. Note that potentially the power-law could flatten
towards smaller mass scales although current simulations show no hits
of such a behavior and in addition we are approaching the asymptotic
behavior in the power spectrum on these scales(CHECK). For DM halos
more (less) massive than the MW we expect a larger (smaller) boost
from substructures, simply because of the larger (smaller) mass range
down to the minimum mass $M_\rmn{lim}$. We capture this halo mass
dependence with the second normalization in Eq.~(\ref{eq:LSub}):
\begin{equation}
C_2(\mvir)=\left(\frac{\mvir}{M_{\rm 200sim}}\right)^{0.226}\,,
 \label{eq:C2}
\end{equation}
where $M_{\rm 200sim}=1.9\times10^{12}\msun$ is the mass of the MW
halo in the simulation \cite{2008Natur.456...73S}.

@@

In Fig.~\ref{fig:radial_emis} we show the radius with the maximum
luminosity for substructures and smooth density profiles, both for
three different mass scales. We find that the smooth luminosity per
logarithmic radial bin peaks at $x \dd^2 L/\dd x^2+\dd L/\dd x = 0
\rightarrow r \propto \rvir/3c$. Hence the main contribution to the
luminosity for large mass scales, which are usually associated with a
low concentrated DM halos, is focused to the regime around 10\% of
$\rvir$.  It clearly shows how that the main contribution from the
substructures is coming from outer parts of the DM halos, while the
main contribution from the smooth density profiles peaks around the
scale radius. Not that, even though the bulk of substructures have
been erased in the central regions of DM halos, a cluster in
projection has a significant enhancement due to substructures at a
radius of just a few percent of $\rvir$.
\begin{figure}
  \includegraphics[width=0.99\columnwidth]{figures/emissiv.sub.eps}
  \caption{\it Luminosity dominated radius for different mass
    scales. The solid lines show the emissivity from a cuspy
    $\beta=1.0$ NFW density profile for three different mass scales;
    an $\mvir=10^{15}\,\msun$ galaxy cluster (red), an
    $\mvir=10^{12}\,\msun$ galaxy (orange), and an
    $\mvir=10^{8}\,\msun$ dwarf galaxy (green). The dashed lines show
    the emissivity from substructures for the same three mass
    scales. All emissivities have been normalized with the luminosity
    within $\rvir$ from a cuspy NFW profile. We have assumed the
    standard value for the limiting substructure mass of
    $M_\rmn{lim}=10^{-6}\,\msun$.}
  \label{fig:radial_emis}
\end{figure}

@@

We adopt the spectrally and spatially universal gamma-ray model
developed by Pinzke \& Pfrommer \cite{2010MNRAS.409..449P} to estimate
the emission from decaying $\pi^0$:s that dominates over the inverse
Compton emission from primary and secondary electrons above 100~MeV in
clusters. The gamma-ray formalism was derived from high resolution
simulations of clusters of galaxies that included radiative
hydrodynamics, star formation, supernova feedback, and followed CRp
physics using a novel formulation that trace the most important
injection and loss processes self-consistently while accounting for
the CRp pressure in the equation of motion
\cite{2008A&A...481...33J,2007A&A...473...41E,2006MNRAS.367..113P}.
We note that the overall normalization of the CRp and gamma-ray
distribution scales with the maximum acceleration efficiency at
structure formation shock waves. Following recent observations at
supernova remnants \cite{2009Sci...325..719H} as well as theoretical
studies \cite{2005ApJ...620...44K}, we adopt a realistic value of this
parameter and assume that 50\% of the dissipated energy at strong
shocks is injected into CRps while this efficiency rapidly decreases
for weaker shocks \cite{2007A&A...473...41E}.

\subsection{Detecting Particle Dark Matter}
\label{sect:PF}
Besides the intrinsic interest in the gamma-ray flux from galaxy
clusters generated by conventional hadronic and electromagnetic
process - which should be close to observability with the Fermi-LAT
data \cite{Berezinsky:1996wx,2007A&A...473...41E,2010MNRAS.409..449P}
- the possible contribution from WIMP dark matter is of greatest
interest. A WIMP (Weakly Interactive Massive Particle) which fulfils
the WMAP bounds on the relic density of cold dark matter (CDM) of
\cite{Komatsu:2010fb}
$$\Omega_{CDM}h^2=0.112\pm 0.0056,$$ will in many cases naturally give
a $\gamma$ yield which may be observable. In addition, there are
possible enhancement effects known, such as the astrophysical boost
from dense substructure of dark matter halos, or the particle physics
boost from the Sommerfeld effect, which may increase the chances of
detection further.

@@

{\it IC -- dust and starlight}\\ Galaxy clusters are typically
characterized by the hot gas in the ICM and the collection of
gravitationally bound galaxies. The emission in the IR and UV
wavelengths emerge from dust and starlight in both the galaxies and
the ICM (e.g. \cite{2006ApJ...648L..29P} and
\cite{2009MNRAS.399.1694G}). Much of the radiation leaks out from the
galaxy into the ICM, which give rise to similar spectral distributions
for the galaxies and the ICM in these wavelengths. Hence, we use the
accurately measured light from galactic dust and starlight for the ISM
to model the ICM. We characterize the spectra through a fit to the
galactic spectra presented in \cite{2006ApJ...648L..29P}. The spatial
distribution of the SD light in clusters is extracted from a stacked
emission analysis of Sloan Digital Sky Survey (SDSS) data at the
redshift $\sim 0.25$ \cite{2005MNRAS.358..949Z}. Finally, we fix the
normalization for the light from stars and dust individually in a
galaxy cluster using using the IR to X-ray luminosity scaling relation
found in \cite{2008A&A...490..547G}. The X-ray luminosities for the
clusters used in this work are listed in the extended HIghest X-ray
FLUx Galaxy Cluster Sample (HIFLUGCS) \cite{2002ApJ...567..716R} from
the Rosat all-sky survey. Note that we have added the Virgo cluster to
the sample that we nevertheless refer to as extended HIFLUGCS
catalogue in the following.

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/lum.stars.eps}
\caption{\it accumulative luminosity}
 \label{fig:SD_lum}
\end{figure}
\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/fit.porter.v2.eps}
\caption{\it Spectral dependence of radiation fields in a cluster of
  galaxies. The black dotted line in the left peak show the spectrum
  of CMB photons using a black body with a temperature of
  $2.73$~K. The crosses in the middle and right peaks represent the
  measured spectra from stars and dust (SD), respectively, and are
  derived in \cite{2006ApJ...648L..29P} for a galaxy. We normalize the
  individual SD spectrum separately using the observed luminosity from
  SD in clusters. The SD luminosity is related to the cluster mass
  through Eq, where we use have used the mass
  $\mvir=4.0\times10^{13}\msun$ in this figure. Finally we renormalize
  the SD spectra to the radius $r=0.071\rvir$, where the smooth energy
  density of the SD light (see Fig.~\ref{fig:SD_Edens}) equals the
  energy density of the CMB. The red dashed lines show the fitted SD
  spectral model.}
 \label{fig:SD_spectra}
\end{figure}
\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/ucool.eps}
 \caption{\it Spatial dependence of the energy density of radiation
   fields in a cluster of galaxies. The energy density of the CMB,
   shown by the blue dotted line, is isotropic throughout the cluster,
   hence represented by a flat profile with
   $u_\rmn{cmb}=0.26\,\rmn{eV}\,\rmn{cm}^{-3}$. The energy density of
   the light from stars and dust (SD) is denoted by the red dashed
   line and the solid purple line for the total SD light and the
   smooth SD light, respectively. The SD light has been renormalized
   to a cluster with the mass $\mvir=4.0\times10^{13}\msun$. Finally we
   show the energy density two magnetic field models with a central
   magnetic field of 3~$\mu G$. The magnetic field scales with the gas
   density to the power $\alpha_\rmn{B}$; green dash-dotted line
   ($\alpha_\rmn{B}=0.5$) and orange dash-dotted line
   ($\alpha_\rmn{B}=0.7$). Note that the SD radiation is dominating
   the energy density inside $\sim0.1\,\rvir$.}
 \label{fig:SD_Edens}
\end{figure}

\section{Spectral gamma-ray distribution}
\label{sect:spectral}
Spectrally resolved indirect DM searches have the advantage of probing
different DM models through their characteristic spectral
distributions. To make current and future DM searches more effective
it is important to know in which energy band to focus the efforts in
order to maximize potential DM signal over expected background.

@@

In Fig.~\ref{fig:IR_comp} we show the total IC emission as well as its
individual contributions from different IC upscattered radiation
fields. The left panel show the gamma-ray emission from the LP model
and the right panel from the $\Kp$ BM model, and for comparison we
overplot the emission expected from the CRps. Due to the flat electron
and positron spectrum resulting from the LP DM model and the smaller
mean energy of the CMB compared to the SD, we find that the
upscattered CMB photons dominate the total DM IC emission in the
energy regime below 100~GeV, while the SD dominate above this
energy. For the BM models this transition is shifted towards a smaller
energy since the electron and positron spectrum has a steep spectral
distribution that suppresses low energy photon fields. This
suppression is also seen in the relative normalization of the flux
where the IC from both CMB and dust photons are lower compared to the
starlight simply because the starlight has a higher energy, hence it
is upscattered by more abundant low energy electrons and positrons
(c.f. Fig.~\ref{fig:q_DM}). At the highest gamma-ray energies of about
100~GeV and above, the IC from starlight steepens because it both
probes the high energy tail of the electrons as well as suffer from
the KN suppression. Even though the SD component seems to be
suppressed compared to the CMB at energies around 100~MeV, it has a
relative large impact on the steady state electron spectrum. If we
remove the SD cooling, we find a factor two larger IC flux at low
energies for a cluster of the size of Fornax. For more massive
clusters we expect a larger IC flux from the SD component, but a
smaller contribution from the CMB because of the more effective
cooling.

>>>>

%NEW
\newcommand{\rmn}{\mathrm}
\newcommand{\bx}{{\bf x}}
\newcommand{\fg}{{F_\gamma}}
\newcommand{\sg}{{S_\gamma}}
\newcommand{\psf}{\theta_\rmn{res}}
\newcommand{\ph}{\rmn{ph}}
\newcommand{\eph}{E_\ph}
\newcommand{\nph}{N_\ph}
\newcommand{\vir}{\rmn{vir}}
\newcommand{\gal}{\rmn{gal}}
\newcommand{\sd}{\rmn{SD}}
\newcommand{\clu}{\rmn{cl}}
\newcommand{\sfe}{\rmn{sfe}}
\newcommand{\sub}{\rmn{sub}}
\newcommand{\msun}{M_\odot}
\newcommand{\ee}{E_\rmn{e}}
\newcommand{\stars}{\rmn{stars}}
\newcommand{\dust}{\rmn{dust}}
\newcommand{\lx}{L_\rmn{X}}
\newcommand{\s}{\rmn{s}}
\newcommand{\Kp}{\rmn{K}^\prime}
\newcommand{\Ip}{\rmn{I}^\prime}
\newcommand{\Js}{\rmn{J}^*}
\newcommand{\Jp}{\rmn{J}^\prime}
\newcommand{\B}{\rmn{B}}
\newcommand{\bsub}{\B_\rmn{sub}}
\newcommand{\qCR}{q_{\rmn{CR}-\ensuremath{\pi^0}}}
\newcommand{\ds}{{\sc DarkSUSY}}
\newcommand{\mlim}{M_\rmn{lim}}

@@

\begin{abstract}
... to be written...  if $\mlim \approx 10^{-6}\,\msun$ then the boost
  from SFE is smaller than about a factor 5 in the MW and 2 in Fornax, or
  alternatively if SFE corrext then $\mlim > 10^2\,\msun$.
\end{abstract}

@@

This work studies the prospects for detecting extended gamma-ray
emission from galaxy clusters induced by both DM as well as CR
protons. In section two we discuss the theory of DM and gamma-rays. In
particular, we focus one the leptophilic and supersymmetric DM models
used in this work, and outline the framework for estimating the
gamma-ray emission from various radiative processes. In section three
we calculate the gamma-ray fluxes and spectral distributions from four
clusters and compare it to upper limits set by Fermi-LAT. For the same
clusters we derive the gamma-ray surface brightness profiles in
section four. To extend the analysis to more clusters and find
promesing targets we estimate the fluxes from DM and CRs of all
clusters in the extended HIFLUGCS sample in section five. In section
six we conclude and discuss our results.

% --- section:  --- %
\section{Theory}
\label{sect:theory}
We start the section by discussing two different, but well motivated DM
models; leptophilic and supersymmetric DM. In addition we present the
gamma-ray framework that is used to calculate the emission for these DM
models using an Einasto DM density profile and the expected
enhancement from DM substructures. In the end of this sections we
outline the framework of inverse Compton emission on both CMB photons
as well as photons from stars and dust.

\subsection{Detecting Particle Dark Matter}
\label{sect:PF}
Besides the intrinsic interest in the gamma-ray flux from galaxy
clusters generated by conventional hadronic and electromagnetic
process - which should be close to observability with the Fermi-LAT
data \cite{1997ApJ...487..529B,2007A&A...473...41E,2010MNRAS.409..449P}
- the possible contribution from WIMP dark matter is of greatest
interest. A WIMP (Weakly Interactive Massive Particle) which fulfils
the WMAP bounds on the relic density of cold dark matter (CDM) of
\cite{Komatsu:2010fb}
$$\Omega_{CDM}h^2=0.112\pm 0.0056,$$ will in many cases naturally give
a $\gamma$ yield which may be observable. In addition, there are
possible enhancement effects known, such as the astrophysical boost
from dense substructure of dark matter halos, or the particle physics
boost from the Sommerfeld effect, which may increase the chances of
detection further.

@@

\frac{\dd \sg}{\dd \eg}(\psi,>\eg) = \int_{\Delta\Omega} \dd\Omega \int_{\rm los}

@@

radiation (BM-Cont), and CR proton induced $\pi^0$ that decays into

@@

is given by:
\begin{eqnarray}
\B_{\rmn{tot},i}(r,\sigv) = \left\{\begin{array}{cc}
\B_\sfe(\sigv)\,\B_{\sub,i}(r) &\rmn{for\,\,LP}\\
\B_{\sub,i}(r) &\rmn{for\,\,BM\,.}\end{array}\right.
\end{eqnarray}
It is the product of enhancement factors from SFE $\B_\sfe(\sigv)$
(see Sect.~\ref{sect:LP}) and from substructure enhancement over the
smooth halo contribution $\B_{\sub,i}(r) =
1+\rho_{\sub,i}^2(r)/\rho^2(r)$ (see Sect.~\ref{sect:subst}). The DM
source function from the smooth halo for each process is written in
the form:

@@

Here runs the subscript $j$ over all kinematically allowed gamma-ray
producing channels, each with the spectrum $\frac{\dd

@@

profile provides a good fit to both the observed and simulated

@@

should also be noted that a recent observations of the Abell 383
galaxy cluster finds a core like density profile with an inner slope
of $\beta=0.6$, where $\beta\gtrsim 1$ can be ruled out to a $>95$\%
confidence \cite{2011ApJ...728L..39N}. These observations are based on
lensing and X-ray measurements as well as the stellar velocity
dispersion of the central galaxy. In Eq.~(\ref{eq:dens_ein}) we denote
the density where the profile has a slope of $-2$ by $\rho_{-2}$, and
the radius by $r_{-2}$.  Assuming that all the flux from an NFW profile

@@

that $\rho_{-2} = \rhos/4$. [MIGHT REMOVE THIS, COULD NOT REPRODUCE IT, CAN
MOTIVATE THE FACTOR 4 FROM MATCHING THE MASSES FROM THE EINSASTO WITH
THE NFW INSTEAD] SHOW INTEGRATION, COMPUTE RMAX, DISCUSS
FIG 1+2 SMOOTH COMPONENT

@@

\subsubsection{Substructures}
\label{sect:subst}
High-resolution DM only simulations of Milky Way (MW) type halos find
substantial amount of substructures in the periphery of DM halos,
while the substructures in the center suffer from dynamical friction
and tidal effects depleating their central number densities. Since
the rate of which DM is annihilating depends on the density squared,
the resulting flux from substructures is boosted compared to the
smooth density distribution. Recent high resolution simulations
suggest a flux enhancement from substructures of the order of ten up
to a few 100 \cite{2008MNRAS.391.1685S, 2008JPhCS.125a2008K} for a MW
size halo. We use a double power-law function to fit \footnote{Our
  approach of fitting the scaling behavior of $\rmn{L_{sub}}(<r)$
  directly from numerical simulations self-consistently accounts for
  the radial dependence of the substructure concentration
  \cite{2008MNRAS.391.1685S}.} the luminosity from the smooth
component of substructures (i.e. substructures within substructures
are not included) inside radius $r$ which is determined for the
Aquarius simulations
\cite{2008MNRAS.391.1685S,2008Natur.456...73S}. Our best fit is given
by
\begin{eqnarray}
  {\rm L_{sub}}(<x) &=& a_0\,C(\mvir)\,{\rm L_{200sm}}(\mvir)\,x^{a_1
    \displaystyle{x}^{a_2}}\,,\quad\rmn{where} \nonumber\\
   a_0 &=& 0.76\,,\quad a_1 = 0.95\,,\quad  a_2 = -0.27\,, \quad 
\rmn{and} \nonumber\\
\label{eq:LSub} \\
  C(\mvir) &=& 0.023\,\left(\frac{\mvir}{M_{\rm lim}}\right)^{0.226}\,.
\label{eq:Csub}
\end{eqnarray}
Here $a_i$ denote our fitted variables, ${\rm L_{200sm}}$ is the
luminosity from the smooth halo without substructures within $\rvir$
and $x= \left(\frac{r}{\rvir}\right)$. The theoretical smallest mass
of subhalos $M_{\rm lim}$ is determined from free streaming length of
DM (CHECK), where we assume that the mass power-law scaling relation
in Eq.~(\ref{eq:Csub}) is valid down to the free streaming mass of DM
halos. This is conventionally in the cold dark matter universe
\cite{2001PhRvD..64h3507H,2005JCAP...08..003G} taken to be
$10^{-6}\msun$. Note that potentially the power-law could flatten
towards smaller mass scales although current simulations show no hits
of such a behavior and in addition we are approaching the asymptotic
behavior in the power spectrum on these scales(CHECK). We derive the
normalization function $C(\mvir)$ in Eq.~(\ref{eq:Csub}) from the
simulations in \cite{2008Natur.456...73S}:
\begin{equation}
\left(\frac{\mvir\,M_{\rm res,sim}}
{M_{\rm 200sim}\,M_{\rm lim}}\right)^{0.226}\,.
 \label{eq:C2}
\end{equation}
Here $M_{\rm 200sim}=1.9\times10^{12}\msun$ is the mass of the MW halo
in the simulation and $M_{\rm res,sim}=10^5\msun$ is the mass of the
smallest resolved subhalos in the MW simulation. For DM halos more
(less) massive than the MW we expect a larger (smaller) boost from
substructures, simply because of the larger (smaller) mass range down
to the minimum mass $M_\rmn{lim}$.

Finally we derive the squared density profile for the substructures
using
\begin{eqnarray}
\rho_\rmn{sub}^2(r) &=& \frac{\dd \rmn{L_{sub}}}{\dd V} \frac{1}{\rmn{L_{200sm}}}
\int_{\rvir} \dd r' 4\pi r'^2 \rho^2(r')\qquad\rmn{where}\nonumber\\
 \frac{\dd \rmn{L_{sub}}}{\dd V} &=& a_0\,a_1\,C(\mvir)\,x^{a_2+a_1\,x^{a_2}}\,
\left[\frac{1+a_2\log(x)}{4\pi\,r^3}\right]\quad\rmn{and}\nonumber\\
x &=& r/\rvir\,.
\end{eqnarray}

@@

In Fig.~\ref{fig:radial_emis} we show the radius with the maximum
luminosity for substructures and smooth density profiles, both for
three different mass scales. We find that the smooth luminosity per
logarithmic radial bin peaks at $x \dd^2 L/\dd x^2+\dd L/\dd x = 0
\rightarrow r \propto \rvir/3c$. Hence the main contribution to the
luminosity for large mass scales, which are usually associated with a
low concentrated DM halos, is focused to the regime around 10\% of
$\rvir$.  It clearly shows how that the main contribution from the
substructures is coming from outer parts of the DM halos, while the
main contribution from the smooth density profiles peaks around the
scale radius. Not that, even though the bulk of substructures have
been erased in the central regions of DM halos, a cluster in
projection has a significant enhancement due to substructures at a
radius of just a few percent of $\rvir$.
\begin{figure}
  \includegraphics[width=0.99\columnwidth]{figures/emissiv.sub.eps}
  \caption{\it Luminosity dominated radius for different mass
    scales. We show the normalized luminosity per logaritmic radial
    bin which corresponds to $x^3\rho^2/\rho^2_{200}$ where
    $\rho_{200}$ is the density at $\rvir$. The solid lines represents
    cuspy $\beta=1.0$ NFW density profiles for three different mass
    scales; an $\mvir=10^{15}\,\msun$ galaxy cluster (red), an
    $\mvir=10^{12}\,\msun$ galaxy (orange), and an
    $\mvir=10^{8}\,\msun$ dwarf galaxy (green). The dashed lines show
    the contribution from substructures for the same three mass
    scales. All luminosities have been normalized with the luminosity
    within $\rvir$ from a cuspy NFW profile. We have assumed the
    standard value for the limiting substructure mass of
    $M_\rmn{lim}=10^{-6}\,\msun$. The peaks show that the flux is not
    dominated by the center of the halos, but instead by the core
    region for the smooth density profile and by the outer parts for
    the substructures.}
  \label{fig:radial_emis}
\end{figure}

@@

We adopt the universal spectral and spatial gamma-ray model developed
by Pinzke \& Pfrommer \cite{2010MNRAS.409..449P} to estimate the
emission from decaying $\pi^0$:s that dominates over the inverse
Compton emission from primary and secondary electrons above 100~MeV in
clusters. The gamma-ray formalism was derived from high resolution
simulations of clusters of galaxies that included radiative
hydrodynamics, star formation, supernova feedback, and followed CRp
physics using a novel formulation that trace the most important
injection and loss processes self-consistently while accounting for
the CRp pressure in the equation of motion
\cite{2008A&A...481...33J,2007A&A...473...41E,2006MNRAS.367..113P}.
We note that the overall normalization of the CRp and gamma-ray
distribution scales with the maximum acceleration efficiency at
structure formation shock waves. Following recent observations at
supernova remnants \cite{2009Sci...325..719H} as well as theoretical
studies \cite{2005ApJ...620...44K}, we adopt a realistic value of this
parameter and assume that 50\% of the dissipated energy at strong
shocks is injected into CRps while this efficiency rapidly decreases
for weaker shocks \cite{2007A&A...473...41E}.

@@

{\it IC -- dust and starlight}\\ Galaxy clusters are typically
characterized by the hot gas in the ICM and the collection of
gravitationally bound galaxies. The emission in the IR and UV
wavelengths emerge from dust and starlight in both the galaxies and
the ICM (e.g. \cite{2006ApJ...648L..29P} and
\cite{2009MNRAS.399.1694G}). Much of the radiation leaks out from the
galaxy into the ICM, which give rise to similar spectral distributions
for the galaxies and the ICM in these wavelengths. Hence, we use the
accurately measured light from galactic dust and starlight in the ISM
to model the ICM. In Fig.~\ref{fig:SD_spectra} we characterize the
spectral shape through a fit to the galactic spectra presented in
\cite{2006ApJ...648L..29P}. We fix the normalization for the energy
density of light from stars and dust in a galaxy cluster individually
using observed luminosities from SD. In addition, we attach a mass
scaling to the normalization which we adopt from a simple
mass-to-starlight luminosity scaling relation derived from
observations of the brightest cluster galaxy
\cite{2010ApJ...713.1037H}. The figure is showing the spectra at
$r=0.03\rvir$, which is the radius there the SD energy density equals
the energy density of the CMB black-body distribution in a galaxy
cluster. Inside this central radius the SD component is dominating,
which is shown in Fig.~\ref{fig:SD_Edens} where we compare the energy
densities from different radiation fields in a galaxy cluster with the
mass $\mvir=10^{15}\msun$. For this figure we use two different
profiles for the SD energy density, where the total profile includes
the contribution from the ICL, the BCG and all the galaxies, while the
galaxies are excluded in the smooth profile. We find that even for a
cluster with a relative small mass the energy density of the SD
components dominates over the CMB and the magnetic fields (with a
central B field of 3~$\mu G$) within about 10\% of $\rvir$. Outside
this radius the CMB is dominating the energy density of the cluster,
hence the magnetic fields are always subdominant in the cluster and
can safely be removed from cluster IC calculations for a reasonable
central B-field and spectral index. Note that in this work we keep the
contribution from B-fields to the total energy density for
consistency. Also note that we extract the spatial distribution of the
SD light in clusters from a stacked emission analysis of Sloan Digital
Sky Survey (SDSS) data at the redshift $\sim 0.25$
\cite{2005MNRAS.358..949Z} (c.f. Fig.~\ref{fig:SD_spatial}).

ADD DISCUSSION ABOUT ACCUMILATIVE LUMINOSITY

\del{
found the IR to X-ray luminosity scaling relation found in
\cite{2008A&A...490..547G}. The X-ray luminosities for the clusters
used in this work are listed in the extended HIghest X-ray FLUx Galaxy
Cluster Sample (HIFLUGCS) \cite{2002ApJ...567..716R} from the Rosat
all-sky survey. Note that we have added the Virgo cluster to the
sample that we nevertheless refer to as extended HIFLUGCS catalogue in
the following.
}

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/fit.porter.v2.eps}
\caption{\it Spectral dependence of radiation fields in a cluster of
  galaxies. The black dotted line in the left peak show the spectrum
  of CMB photons using a black body with a temperature of
  $2.73$~K. The crosses in the middle and right peaks represent the
  measured spectra from stars and dust (SD), respectively, and are
  derived in \cite{2006ApJ...648L..29P} for a galaxy. We normalize the
  individual SD spectrum separately using the observed luminosity from
  SD in clusters. The SD luminosity is related to the cluster mass
  through Eqns.~(\ref{eq:N_stars}-\ref{eq:u_SD}), where we use have
  used the mass $\mvir=10^{15}\msun$ in this figure. Finally we
  renormalize the SD spectra to the radius $r=0.03\rvir$, where the
  smooth energy density of the SD light (see Fig.~\ref{fig:SD_Edens})
  equals the energy density of the CMB. The red dashed lines show the
  fitted SD spectral model.}
 \label{fig:SD_spectra}
\end{figure}
\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/ucool.eps}
 \caption{\it Spatial dependence of the energy density of radiation
   fields in a cluster of galaxies. The energy density of the CMB,
   shown by the blue dotted line, is isotropic throughout the cluster,
   hence represented by a flat profile with
   $u_\rmn{cmb}=0.26\,\rmn{eV}\,\rmn{cm}^{-3}$. The energy density of
   the light from stars and dust (SD) is denoted by the red dashed
   line and the solid purple line for the total SD light and the
   smooth SD light, respectively. For comparison we show the energy
   density of the stars and a low dust model (SLD) with the light blue
   line. The SD light has been renormalized to a cluster with the mass
   $\mvir=10^{15}\msun$. Finally we show the energy density two
   magnetic field models with a central magnetic field of 3~$\mu
   G$. The magnetic field scales with the gas density to the power
   $\alpha_\rmn{B}$; green dash-dotted line ($\alpha_\rmn{B}=0.5$) and
   orange dash-dotted line ($\alpha_\rmn{B}=0.7$). Note that the SD
   radiation is dominating the energy density inside
   $\sim0.1\,\rvir$.}
 \label{fig:SD_Edens}
\end{figure}

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/lum.stars.eps}
\caption{\it Accumulative luminosity as a function of radius. We show
  the luminosity without substructures inside radius $r$. The solid
  line show the emission from stars and dust (SD) and the IC
  upscattered SD light is shown by the dashed lines. The red lines
  show the total emission including the brightest cluster galaxy, the
  intra cluster light and the additional intra cluster galaxies, while
  these galaxies have been cut out in the smooth component shown by
  the orange lines. We normalize the SD light and the IC upscattered
  SD light with the luminosity from the smooth component within
  $\rvir$ for each component. We find that the gamma-ray IC
  luminosities are dominated by the central regime, while the SD light
  is mainly build up in the outer parts of the cluster. In addition we
  show that the SD light trace the cuspy NFW density profile (blue)
  outside 10\% of $\rvir$.}
 \label{fig:SD_lum}
\end{figure}

\section{Spectral gamma-ray distribution}
\label{sect:spectral}
Spectrally resolved indirect DM searches have the advantage of probing
different DM models through their characteristic spectral
distributions. To make current and future DM searches more effective
it is important to know in which energy band to focus the efforts in
order to maximize potential DM signals over expected background.

@@

One of the main contributions to the expected extended gamma-ray is
coming from the IC upscattered photons on various photon background
fields. Depending on the spatial and spectral distribution of
electrons and positrons as well as the photon background field, the
resulting spectral distributions of gamma-ray can differ
greatly. Hence it is interesting to understand which spectral features
and energy regimes that are dominating for the different DM models.

If the substructure boost in clusters is significant, then the total
electron and positron distribution is dominated by its radial profile
outside the very center where the smooth density profile dominates
over the substructures and the cooling from SD dominates over the
CMB. Especially since the SD distribution is centrally peaked and the
substructure distribution peaks in the outskirts around $\rvir$, the
overlap is small. This results in a suppression of the IC upscattered
SD photons relative to the IC upscattered CMB, the FSR and the
continuum emission. However if substructures are only marginally
dominating over the smooth distribution in a cluster, the SD component
becomes more important, especially in the 100~GeV to 1~TeV energy
regime where the IC on CMB photons is small.
